---
title: "Hold and Save Analysis"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
```{r include=FALSE}
library(gridExtra)
library(ggplot2)
library(data.table)
library(knitr)
library(dplyr)
```

### What are we even doing here

The purpose of this analysis is to research the impact of boosting save and hold
point values for the league.  

_Note: No hitters, perfect games, and pickoffs are being left out here_

```{r include=FALSE}
# generate pitching point value table:
pt_vals <- c(3,-1,5,-3,5,-3,1,-1,-1,-1,3,3)
names(pt_vals) <- c("IP", "ER", "W", "L", "SV", "BSv", "SO", "H", "BB", "HBP",
                    "QS", "Hold")
pt_edit <- c(3,-1,5,-3,7,-3,1,-1,-1,-1,3,5)
names(pt_edit) <- names(pt_vals)
```

#### Current pitching point values

```{r echo=FALSE,  results = 'asis'}
kable(pt_vals)
```

#### Proposed values
```{r echo = FALSE,  results = 'asis'}
kable(pt_edit)
```

```{r include=FALSE}
# tables acquired from bbref
# export as csv -> copy/paste into sublime (text editor) and save to wd
std_pitching <- fread("standard_pitching.csv")
start_pitching <- fread("starting_pitching.csv")
rel_pitching <- fread("relief_pitching.csv")
```

### Data
Here's a quick look at each of the tables we're pulling in: standard pitching,
relief pitching, and starting pitching.

Standard pitching is our source of IP, W-L, SO, H, ER, SV, BB, K, and HBP.
Unfortunately it's missing two stats that we have to get from the other tables, 
respectively: Holds and QS.

#### Standard pitching

```{r echo=TRUE}
#show sample data
std_pitching[1:10,]
```

#### Relief pitching

```{r}
rel_pitching[1:10,]
```

#### Starting pitching

One thing you'll notice in the starting pitching table is that there are three
lines for Matt Andriese - he was traded midseason. The only row we care about is
the total row, so we'll filter for those later.

```{r}
start_pitching[1:10,]
```

#### Cleaning the data and putting it all together

As mentioned previously, we need to find all the players that have been traded 
so that we don't end up with multiple records for the same player. The standard
pitching table has everybody included in it. Conveniently, it also always lists 
the "Total" row first, so we can use a trick to get just the first row for each
name. We'll need to do this for all of our data tables.

```{r}
tot_all_pitchers <- std_pitching[, .SD[1], by = Name]
tot_rel_pitchers <- rel_pitching[, .SD[1], by = Name]
tot_start_pitchers <- start_pitching[, .SD[1], by = Name]
```

Next we'll need to filter our relief and starting pitcher tables down to just 
the columns that we're intersted in, Holds and QS, respectively.

```{r}
hold_rel_pitchers <- tot_rel_pitchers[, .(Name, Hold, BSv)]
qs_start_pitchers <- tot_start_pitchers[, .(Name, QS)]
```

Lastly we set keys for the data tables and join them all together:
```{r}
setkey(tot_all_pitchers, "Name")
setkey(hold_rel_pitchers, "Name")
setkey(qs_start_pitchers, "Name")

full_pitchers <- hold_rel_pitchers[qs_start_pitchers[tot_all_pitchers]]
```

Here's what our new data table looks like:
```{r}
full_pitchers[1:10,]
```

Now we need to clean up the NA's, since that'll cause us problems when we try to
generate point totals. We can infer that these should be 0's if they don't exist
.

```{r}
# Thanks Matt Dowle
# https://stackoverflow.com/questions/7235657/fastest-way-to-replace-nas-in-a-large-data-table
# plaigarism is bad

cleandt <- function(DT) {
for (j in seq_len(ncol(DT)))
  set(DT, which(is.na(DT[[j]])),j,0)
}

cleandt(full_pitchers)
```

The purpose of this is to compare how much better relievers get, so we need to 
devise a way to make sure that we track what position a player falls into.

```{r}
# If a pitcher is not in starting pitcher table label as a reliever
# If a pitcher is in starting pitching and in relieving, label as both
# Otherwise label as starter
full_pitchers[, pos:= ifelse(!Name %in% qs_start_pitchers$Name, "RP",
                              ifelse(Name %in% hold_rel_pitchers$Name, "RP/SP",
                                     "SP"))]
full_pitchers
```

### Analysis

Next up we need to generate point totals for both scoring systems. 

We'll do this by taking the columns that line up with our scoring system, 
turning our data table into a matrix, and then multiplying it by our score
vector. It's a neat trick.

```{r}

# prepoints are scores prior to the scoring change
full_pitchers[, prepoints := 
                full_pitchers[, names(pt_vals), with = FALSE] %>%
                as.matrix() %*% pt_vals]

# postpoints are scores post scoring change
full_pitchers[, postpoints := 
                full_pitchers[, names(pt_vals), with = FALSE] %>%
                as.matrix() %*% pt_edit]

# point change is difference
full_pitchers[, ptchange := postpoints - prepoints]
```

#### Comparing before and after

With scores calculated, what's left is to assess how things changed.

First we'll take a look at how the composition of the top pitchers ranked.

For the purpose of tracking fringe pitchers, we'll look at the top 120 before
and after (12 pitchers/team).

```{r include=FALSE}
full_pitchers[, prerank := frankv(full_pitchers[["prepoints"]], order = -1)]
full_pitchers[, postrank := frankv(full_pitchers[["postpoints"]], order = -1)]
full_pitchers[, rankchange := prerank - postrank]
top_pre_pitchers <- full_pitchers[prerank <= 120]
top_post_pitchers <- full_pitchers[postrank <= 120]
```

```{r echo=FALSE}
top_pre_pitchers$pos %>% table() %>% kable(caption = "Pre-scoring change
                                           breakdown by pitcher type")
```


```{r echo=FALSE}
top_post_pitchers$pos %>% table() %>% kable(caption = "Post-scoring change
                                           breakdown by pitcher type")
```

Here are the interesting findings: players that enter and leave the top 120.

The relievers that enter the top 120:

```{r echo=FALSE}
top_post_pitchers[!Name %in% top_pre_pitchers$Name]
```

And the other pitchers that leave:
```{r echo=FALSE}
top_pre_pitchers[!Name %in% top_post_pitchers$Name]
```

Within the new top 120, it's also worth looking at where the pitchers are by
percentile.

#### Graph time

```{r echo=FALSE}
preg <-
  top_pre_pitchers %>%
    ggplot(aes(x = postpoints, color = pos)) +
    geom_density() +
    ggtitle("Point distribution before scoring change")

postg <-
  top_post_pitchers %>%
    ggplot(aes(x = postpoints, color = pos)) +
    geom_density() +
    ggtitle("Point distribution after scoring change")

preg
postg
```


```{r echo=FALSE}
top_pre_pitchers %>%
  ggplot(aes(x = factor(pos), y=prepoints)) +
  geom_boxplot() +
  ggtitle("Point distribution within pitcher types pre-scoring change")

top_post_pitchers %>%
  ggplot(aes(x = factor(pos), y=postpoints)) +
  geom_boxplot() +
  ggtitle("Point distribution within pitcher types post-scoring change")
```

As shown in the plots above, relievers gain quite a bit in points here:
Before the change, here's how the top relievers points broke down:
```{r echo=FALSE}
top_pre_pitchers[pos != "SP"]$prepoints %>% summary()
```
_This does include dual type pitchers, as they likely will have RP eligibility._
_This causes some problems because Trevor Bauer is one of these._

And here is the after:
```{r echo=FALSE}
top_post_pitchers[pos != "SP"]$postpoints %>% summary()
```

### Conclusion

Adding points for holds and saves is probably a good idea. It should make
relievers comparable with many of the streamers/fringe starters as well as move
some of the elite relievers further up into the pitcher pool.

Importantly, this shouldn't disturb the balance of pitcher/hitter value by much.

Prior to any scoring change, and assuming the top 120 pitchers are rostered (and
perfectly average, so yeah it's a lot of assumptions but come on what else can
we do), each team nets roughly 213 points per week from pitching:
```{r}
#sum all points and divide by 25 weeks * 10 teams
top_pre_pitchers$prepoints %>% sum() / (25*10)
```
With the change, teams would on average gain about 7 points per week:
```{r}
#do the same thing as before
top_post_pitchers$postpoints %>% sum() / (25*10)
```

This seems like a pretty reasonable update to make. If you would like to look at
the data yourself, I have the csv files used to generate this data. 

